Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-30 doc/ApTranscriptome_manuscript.Rmd
Incorporating comments from RD/AE/SCH/AN into Introduction

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009]. At the population level, thermal tolerance has been assessed in dozens of studies often demonstrating adaptive differentiation [@hoffmann2013]. In combination with environmental data, this thermal tolerance data has provided the backbone to dozens of studies examining limits to species geographic ranges [@angert2011; @diamond2013; @gouveia2014], and the impacts of climate change on population viability [@deutsch2008; @kingsolver2013]. At the organismal level, studies of thermal tolerance have largely focused on candidate genes such as heat shock proteins [@dahlgaard1998]. However, estimates of the standing genetic variation [@krebs1997; @krebs1998; @williams2012], studies of differential gene expression [@leemans2000; @sorensen2005; @teets2012], quantitative trait loci (QTL) [@morgan2006] and knock-outs [@takahashi2011] in the model system *Drosophila melanogaster* all implicate dozens of additional genes underlying thermal tolerance. Little is known about the diversity of these genes or their expression patterns, especially in non-model species.

The study of whole organism gene expression (e.g. transcriptomics) has rapidly increased due to improved RNA sequencing (RNAseq) technology. To date, RNAseq studies have largely focused on differential expression across few environments [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014]. An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014]. Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression.

For evolutionary ecology studies, the goal of RNAseq studies is often to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @oneil2014]. However, the results of RNAseq provides, lists of differentially-expressed genes, are prone to false positives [@gonzalez2013], miss genes with small expression changes that have large phenotypic effects [] and may not be directly informative about organismal status [@feder2005]. Genome-wide association (GWA) studies have faced a similar problem where the phenotypic effect of the few statistically-significant genes are over-estimated while many causal genes of small effect are missed [@beavis1994; @larsson2013]. Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues. Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010]. Analogously, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient (e.g. the thermal reactionome). Instead of focusing on individual genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively.

@@

Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed the expression patterns of expression for genes involved in thermal tolerance in these species. Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.

>>>>

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@cowles1939; @kingsolver2009]. The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@deutsch2008; @kingsolver2013; @hoffmann2013] and these differences are often genetic [@krebs1997a; @kellermann2012]. To understand how this diversity of thermal tolerance evolved, we need more information on the genetic basis of thermal tolerance. Studies of the genetic basis of thermal tolerance have largely focused on candidate genes such as heat shock proteins [@dahlgaard1998]. However, estimates of the standing genetic variation [@krebs1997; @krebs1998; @williams2012], studies of differential gene expression [@leemans2000; @sorensen2005; @teets2012], quantitative trait loci (QTL) [@morgan2006] and knock-outs [@takahashi2011] in the model system *Drosophila melanogaster* all implicate dozens of additional genes underlying thermal tolerance. Little is known about the diversity of these genes or their expression patterns, especially in non-model terrestrial species.

The study of whole organism gene expression (e.g. transcriptomics) has increased in tandem with improved RNA sequencing (RNAseq) technology. In evolutionary ecology, RNAseq studies to date have largely focused on differential expression across few environments [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014] and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014]. Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies that look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is.

Moreover, the results provided by RNAseq data do not always directly address the aims of studies. For evolutionary ecology studies, the goal of RNAseq studies often is to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @oneil2014]. However, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], miss genes with small expression changes that have large phenotypic effects [@padovan2013] and may not be directly informative about organismal status [@feder2005]. Genome-wide association (GWA) studies have faced a similar problem: phenotypic effects of the few statistically-significant genes are over-estimated while many causal genes of small average effect are missed [@beavis1994; @larsson2013]. Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues. Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010].

Here, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient; the *thermal reactionome*. Instead of focusing on individual genes with differential expression, we compare overall patterns in such reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago [bernice??].

Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species. Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.

## Why ants?

@@
